Extended Data

### 2021-10-21 doc/manuscript/Dow_manuscript_Nature.Rmd
"leaf"--> "canopy foliage"
@camerondow35 , I changed this wording throughout because it gives a more precise and accurate definition.

You might also change the y-axis on Fig. 2 to "Canopy Foliage Stage"

While the responses of leaf phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site *(to our knowledge)*[@dorangeville_peak_2021]. <!-- I'm not aware of any others, but we do want to make sure we're not missing anything. klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
The climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).

@@

Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **125** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **340** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of leaf phenology (measured at ecosystem level via remote sensing).

@@

Both leaf phenology and stem growth milestones occurred earlier at SCBI than at Harvard Forest, with stem growth milestones achieved on average 6-10 days earlier at SCBI (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig. `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`). 
Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).

![**Figure 2 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level leaf phenology, obtained from the MODIS Global Vegetation Phenology product (2001-2018), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003). Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.](tables_figures/DOYtiming_allyears.png){ width=25% }

Both leaf phenology and stem growth were sensitive to spring temperatures (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-5), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan. 1 and $DOY_{25}$ for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window; Fig. `r schematic_summary`). 
The strongest relationships, including but not limited to the critical temperature window (`r  EDT_phenology_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).

For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the critical temperature window resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of leaf greenup advanced with increasing $T_{max}$ during the critical temperature window (p= 0.001 at SCBI; p= 0.1 at HF). 
Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`, `r EDT_phenology_parameters`).
Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs. `r schematic_summary`, `r DOY_timing`; `r EDF_pheno_all`).
Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 1.1-1.9 days/ $^\circ$C for ring-porous species and 3.5-3.6 days/ $^\circ$C for diffuse-porous species at SCBI, and 2.8-7.2 days/ $^\circ$C for ring-porous species and 6.6-7.9 days/ $^\circ$C for diffuse-porous species at Harvard Forest (`r EDT_phenology_parameters`).

Whereas the length of time between leaf greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the critical temperature window compared to those with cooler temperatures (Fig. `r DOY_timing`), there was no consistent lengthening of $L_{PGS}$. 
Specifically, because both $DOY_{25}$ and $DOY_{75}$ advanced in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig. `r schematic_summary`, `r EDF_pheno_all`).

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for leaf development (Fig. `r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020] and NEE [@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak leaf greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senesence[@zani_increased_2020;@keenan_timing_2015].
Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with drought stress[@dorangeville_drought_2018].
Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010]) were uncorrelated within our dendrometer band analysis.

@@

Leaf phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a. MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019]. 
These data were used to determine timing of greenup, midgreenup, senescence, and enhanced vegetation index (EVI2) values for each year. 
(**IAN, I think this could use a bit more detail.**)

>>>>

While the responses of canopy foliage phenology and seasonal CO~2~ sequestration to warming spring temperatures have been relatively well-studied[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999], we know very little about how the longest-lived component of fixed carbon in trees, the woody growth, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site *(to our knowledge)*[@dorangeville_peak_2021]. <!-- I'm not aware of any others, but we do want to make sure we're not missing anything. klesse_integrating_2016 and martin_hydrometeorology_2017 both do this, but for conifers.-->
The climate sensitivity of woody growth phenology and its link to annual growth has never been studied *in-situ* (but see Ref.[@zohner_how_2021] for a controlled sapling experiment).

@@

Using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = **125** trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = **340** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@dorangeville_peak_2021], and set in the context of canopy foliage phenology (measured at ecosystem level via remote sensing).

@@

Both canopy foliage phenology and stem growth milestones occurred earlier at SCBI than at Harvard Forest, with stem growth milestones achieved on average 6-10 days earlier at SCBI (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig. `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`). 
Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).

![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology, obtained from the MODIS Global Vegetation Phenology product (2001-2018), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003). Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.](tables_figures/DOYtiming_allyears.png){ width=25% }

Both canopy foliage and stem growth phenology were sensitive to spring temperatures (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
While the results described below were qualitatively consistent considering both maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) across a range of spring time windows (Extended Data Figures 3-5), we focused our analyses on the strongest relationships: responses to $T_{max}$ over the consecutive weekly time window between Jan. 1 and $DOY_{25}$ for which the correlation between $T_{max}$ and $DOY_{25}$ was strongest (henceforth, critical temperature window; Fig. `r schematic_summary`). 
The strongest relationships, including but not limited to the critical temperature window (`r  EDT_phenology_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$ (`r EDF_climwin`).

For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the critical temperature window resulted in earlier achievement of growth milestones (Figs. `r schematic_summary`, `r EDF_pheno_Tmax_climwin`).
Consistent with findings from previous studies[@parmesan_globally_2003, @ibanez_forecasting_2010,@friedl_tale_2014], the DOY of canopy greenup advanced with increasing $T_{max}$ during the critical temperature window (p= 0.001 at SCBI; p= 0.1 at HF). 
Specifically, greenup DOY advanced 4.5 days/$^\circ$C at SCBI and 2.4 days/ $^\circ$C at Harvard Forest, and mid-greenup also advanced, but to a lesser extent (Fig. `r DOY_timing`, `r EDT_phenology_parameters`).
Similarly, at both sites, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g_{max}}$ all decreased with mean $T_{max}$ during the critical temperature window (Figs. `r schematic_summary`, `r DOY_timing`; `r EDF_pheno_all`).
Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 1.1-1.9 days/ $^\circ$C for ring-porous species and 3.5-3.6 days/ $^\circ$C for diffuse-porous species at SCBI, and 2.8-7.2 days/ $^\circ$C for ring-porous species and 6.6-7.9 days/ $^\circ$C for diffuse-porous species at Harvard Forest (`r EDT_phenology_parameters`).

Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased in years with warmer temperatures during the critical temperature window compared to those with cooler temperatures (Fig. `r DOY_timing`), there was no consistent lengthening of $L_{PGS}$. 
Specifically, because both $DOY_{25}$ and $DOY_{75}$ advanced in years with warm springs, $L_{PGS}$ displayed mixed responses to spring temperatures (Fig. `r schematic_summary`, `r EDF_pheno_all`).

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed phenological advance in the start of stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`)[@keenan_net_2014;@finzi_carbon_2020] and NEE [@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs advance the end to the period of most active stem growth and have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020], is likely driven by cues other than carbon limitation, such as daylength or sink limitation, which also play an important role in autumn canopy foliage senesence[@zani_increased_2020;@keenan_timing_2015].
Rather than growing season length, the primary effect of temperature on annual growth is a negative effect of $T_{max}$ during the peak growing season, which is associated with drought stress[@dorangeville_drought_2018].
Warm springs may also augment summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018]; however, spring temperatures and summer Standardized Precipitation Evapotranspiration Index[@vicente-serrano_multiscalar_2010]) were uncorrelated within our dendrometer band analysis.

@@

canopy foliage phenology data for the years 2001-2018 were extracted for SCBI and Harvard Forest from the MCD12Q2 V6 Land Cover Dynamics product (a.k.a. MODIS Global Vegetation Phenology product)[@friedl_mcd12q2_2019]. 
These data were used to determine timing of greenup, midgreenup, senescence, and enhanced vegetation index (EVI2) values for each year. 
(**IAN, I think this could use a bit more detail.**)
